3.5: Causal Inference and DAGs

### 2019-11-07 static/r/17-r-practice.Rmd
Update class 17, add practice 17

---
title: "3.3 OLS: Model Specification Strategies - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

@@

```{r}
library(corrplot)
bwght %>%
  select(bwght, cigs, faminc, cigprice, motheduc) %>%
  cor(., use="pairwise.complete.obs")
```

## Question 3
Consider the following causal model:

@@

## Question 5
Test each implication given to you by `dagitty.`

@@

```{r}
lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
```

```{r}
lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
```

```{r}
lm(motheduc~cigprice, data = bwght) %>% summary()
```

```{r}
lm(faminc~cigprice, data = bwght) %>% summary()
```

@@

## Question 11
Test these implications appropriately, like you did in Question 5.

```{r}
lm(bwght~cigprice+cigs+motheduc+faminc, data = bwght) %>% summary()
```

```{r}
lm(motheduc~cigprice, data = bwght) %>% summary()
```

Does this model hold up well?

## Question 12
Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

## Question 13
Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

## Question 14
Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.

## Question 15
Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.

>>>>

---
title: "3.5: Causal Inference and DAGs - R Practice"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    #theme: 
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 3

Consider the following causal model:

@@

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 5

Test each implication given to you by `dagitty.`

@@

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

@@

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 11
Test these implications appropriately, like you did in Question 5. Does this model hold up well?

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 12

Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 13

Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 14

Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.

---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

---

## Question 15
Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.
---

<!--ANSWER BELOW HERE-->

```{r}
# PUT CODE HERE
```

### 2019-11-07 static/slides/17-slides.Rmd
Update class 17, add practice 17

h1 <- dagify(profit ~ tech+ip,
             tech ~ ip,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "1")+
  theme_void()+theme(legend.position = "none")

h2 <- dagify(profit ~ tech+ip,
             ip ~ tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "2")+
  theme_void()+theme(legend.position = "none")

h3 <- dagify(profit ~ tech,
             ip ~ tech+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "3")+
  theme_void()+theme(legend.position = "none")

h4 <- dagify(profit ~ ip,
             tech ~ ip+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "4")+
  theme_void()+theme(legend.position = "none")

h5 <- dagify(tech ~ ip+profit,
             ip ~ profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "5")+
  theme_void()+theme(legend.position = "none")

h6 <- dagify(tech ~ profit,
             ip ~ profit+tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE)+
  labs(title = "6")+
  theme_void()+theme(legend.position = "none")

@@

```{r, echo = F}
dagify(wage~educ) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws) %>%
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag(stylized = FALSE)+
  theme_dag_blank()
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```
]

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M+X,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=2 )
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X,
       coords=list(
         x=c(X=1,Y=2),
         y=c(X=1,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A,
       C~B,
       coords=list(
         x=c(X=1,A=1.25, B=1.5, C=1.75, Y=2),
         y=c(X=1,A=1.5,B=1.5,C=1.5,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```
]
---

@@

```{r, echo = F}
dagify(Y~X+B+Z,
       X~A,
       B~A) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A+C,
       coords=list(
         x=c(X=1, A=1, B=1.5, C=2, Y=2),
         y=c(X=1, A=2, B=1.5, C=2, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Hos~Flu+Bus,
       coords=list(
         x=c(Flu=1, Hos=1.5, Bus=2),
         y=c(Flu=1, Hos=2, Bus=1)
       )) %>% 
  ggdag_parents("Hos", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~abil+occup+discr,
       discr~gender,
       occup~gender+discr+abil,
       exposure = "disc",
       outcome = "wage") %>% 
  ggdag_parents("wage", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

1. Ensure that no back-door path remains open
    - Close back-door paths by controlling for a variable along that path
    - Colliders along a path automatically close that path

2. Ensure that no front-door path is closed

@@

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE)+theme_dag_blank()+theme(legend.position = "none")
```

@@

>>>>

h1 <- dagify(profit ~ tech+ip,
             tech ~ ip,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "1")+
  theme_void()+theme(legend.position = "none")

h2 <- dagify(profit ~ tech+ip,
             ip ~ tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "2")+
  theme_void()+theme(legend.position = "none")

h3 <- dagify(profit ~ tech,
             ip ~ tech+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "3")+
  theme_void()+theme(legend.position = "none")

h4 <- dagify(profit ~ ip,
             tech ~ ip+profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "4")+
  theme_void()+theme(legend.position = "none")

h5 <- dagify(tech ~ ip+profit,
             ip ~ profit,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "5")+
  theme_void()+theme(legend.position = "none")

h6 <- dagify(tech ~ profit,
             ip ~ profit+tech,
             coords=list(
               x=c(tech=1, ip=2, profit=3),
               y=c(tech=1, ip=2, profit=1)
               )) %>% ggdag(stylized = FALSE, seed=256)+
  labs(title = "6")+
  theme_void()+theme(legend.position = "none")

@@

```{r, echo = F}
dagify(wage~educ) %>% 
  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws) %>%
  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~educ+conx+year+bckg+loc,
       educ~bckg+year+loc+laws,
       conx~educ,
       bckg~u1,
       loc~u1) %>% 
  ggdag(stylized = FALSE, seed=256)+
  theme_dag_blank()
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```
]

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~X,
                coords=list(
                x=c(X=1,Y=2),
                y=c(X=1,Y=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=1)
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F, fig.retina = 2, fig.align="center"}
dagify(Y~M+X,
               M~X,
                coords=list(
                x=c(X=1,Y=2, M=1.5),
                y=c(X=1,Y=1, M=2 )
              )) %>% 
  tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~Z+X,
       X~Z,
       coords=list(
         x=c(X=1,Y=2, Z=1.5),
         y=c(X=1,Y=1, Z=2)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X,
       coords=list(
         x=c(X=1,Y=2),
         y=c(X=1,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A,
       C~B,
       coords=list(
         x=c(X=1,A=1.25, B=1.5, C=1.75, Y=2),
         y=c(X=1,A=1.5,B=1.5,C=1.5,Y=1)
         )) %>% 
  #tidy_dagitty() %>%
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```
]
---

@@

```{r, echo = F}
dagify(Y~X+B+Z,
       X~A,
       B~A) %>% 
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~X+C,
       X~A,
       B~A+C,
       coords=list(
         x=c(X=1, A=1, B=1.5, C=2, Y=2),
         y=c(X=1, A=2, B=1.5, C=2, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Hos~Flu+Bus,
       coords=list(
         x=c(Flu=1, Hos=1.5, Bus=2),
         y=c(Flu=1, Hos=2, Bus=1)
       )) %>% 
  ggdag_parents("Hos", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(wage~abil+occup+discr,
       discr~gender,
       occup~gender+discr+abil,
       exposure = "disc",
       outcome = "wage") %>% 
  ggdag_parents("wage", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

- Tobacco industry claimed that $cor(smoking, cancer)$ could be spurious due to a confounding `gene` that affects both!
  - Smoking `gene` is unobservable

- Suppose smoking causes `tar` buildup in lungs, which cause `cancer`

- We should *not* control for `tar`, it's on the **front-door path**
  - This is how scientific studies can relate smoking to cancer

```{r, echo = F}
dagify(canc~tar+gene,
       tar~smok,
       smok~gene,
       coords=list(
         x=c(smok=1, tar=1.5, gene=1.5, canc=2),
         y=c(smok=1, tar=1, gene=2, canc=1)
       )) %>% 
  ggdag(stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

1. Ensure that **no back-door path remains open**
    - Close back-door paths by *controlling* for any variable along that path
    - Colliders along a path *automatically* close that path

2. Ensure that **no front-door path is closed**

@@

```{r, echo = F}
dagify(Y~A+B+M,
       M~X,
       X~A+B,
       coords=list(
         x=c(X=1, A=1.5, B=1.5, M=1.5, Y=2),
         y=c(X=1, A=2, B=0, M=1, Y=1)
       )) %>% 
  ggdag_parents("Y", stylized = FALSE, seed=256)+theme_dag_blank()+theme(legend.position = "none")
```

@@

# What are DAGs Good For?

- Omitted variable bias

- Help visualize how later methods (difference-in-difference, fixed effects, regression discontinuity, instrumental variables) help identify causal effectsp
